Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwestern China using organic residue analysis

### 2023-11-14 analysis/paper/paper.Rmd
edit based on comments

To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a]. The new evidence at Guijiabao reveals that millet and rice were both present there in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already been part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period. To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery? Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period? We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.

@@

## Methods
We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

@@

The remaining 95% aliquot was fractioned using solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed using GC-C-IRMS. With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might have come from solvents and reagents. Internal standards were added before and after the lipid extraction for quantification.

@@

```{r tidy-data, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/LY_21_08_20_pot_chrom.CSV"))}
library(tidyverse)

@@

The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)]. The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)). These fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from larger molecules, such as triacylglycerols, due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds had higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005]. However, we may still need to examine their isotopic values to confirm their sources.

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples. A higher yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation. This sample also has a charred interior surface, indicative of a function associated with food processing. Long-chain fatty acids from C20:0 to C32:0 are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high amount of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 have a distribution between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

The results show that the majority of the potsherd samples from the Guijiabao site contain fatty acids from ruminant adipose fats, with one potsherd sample (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A). Analysis of the potsherd from the Gaoshan site (CDG-062) indicates ruminant adipose fats as well. The potsherds (SYG-TN13-E22-1 and SYG-3) from the Guijiabao site dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of these two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for determining food categories.

@@

```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values are plotted against δ13C16:0 values"}
knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
```

>>>>

To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a]. The archaeobotanical evidence at Guijiabao reveals that foxtail millet, broomcorn millet, and rice were present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Huan2022], and millet could have already been part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period. To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery? Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period? We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.

@@

## Methods
We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C~16:0~) and stearic (C~18:0~), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

@@

The remaining 95% aliquot was fractioned using solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed using GC-C-IRMS. With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might have come from solvents and reagents. Internal standards, including FA C~21:0~ and 5-α Cholestane, were added before and after the lipid extraction for quantification.

@@

```{r tidy-data, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/LY_21_08_20_pot_chrom.CSV"))}
library(tidyverse)
library(ggtext)

@@

The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid), C~18:0~ (stearic acid), and monounsaturated fatty acid C~18:1~ (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)). These fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from larger molecules, such as triacylglycerols, due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of medium-chain fatty acids (C~16:0~ and C~18:0~) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds had higher contents of C~18:1~ compared to C~18:0~ and C~16:0~, such as SYG-TN13-E23-3. A higher content of C~18:1~ and C~16:0~ with a lower content of C~18:0~ might indicate the origin of plant oils [@Mayyas2018; @Copley2005]. However, we may still need to examine their isotopic values to confirm their sources.

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples. A higher yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation. This sample also has a charred interior surface, indicative of a function associated with food processing. Long-chain fatty acids from C~20:0~ to C~32:0~ are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high amount of C~16:0~ relative to C~18:0~ might be related to plant waxes [@Post1996; @Copley2005]. However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C~16:0~ and C~18:0~, using GC-C-IRMS. The δ^13^C values of C~16:0~ and C~18:0~ have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C~18:0~ - δ^13^C~16:0~) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C~16:0~ and C~18:0~ have a distribution between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ^13^C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ^13^C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

The results show that the majority of the potsherd samples from the Guijiabao site contain fatty acids from ruminant adipose fats, with one potsherd sample (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A). Analysis of the potsherd from the Gaoshan site (CDG-062) indicates ruminant adipose fats as well. The potsherds (SYG-TN13-E22-1 and SYG-3) from the Guijiabao site dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of these two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021]. Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low ^13^C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for determining food categories.

@@

```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C~16:0~ and C~18:0~. B: The Δ13C (δ^13^C~18:0~-~16:0~) values are plotted against δ^13^C~16:0~ values"}
knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
```
